---
title: 'Module 1: The Basics'
author: "Jasmine Hughes"
date: "9/17/2019"
output: 
  slidy_presentation: default
---


```{r chunksetup, include=FALSE, cache=TRUE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
if(!('Modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('Modules')
gap <- read.csv(file.path('..', 'Data', 'gapminder-FiveYearData.csv'), stringsAsFactors = FALSE)
```

# R as a calculator

```{r computing,echo=TRUE,results='markup'}
2 + 2  # add numbers
2 * 3  # multiply numbers
3 ^ 4  # powers
3 ** 4 # powers
2 * pi # multiply by a constant
```

```{r computing2,echo=TRUE,results='markup'}
sqrt(4^4) # square root
23 %% 2   # remainder
log(10)   # logarithms (what is the default base?)
log(100, base = 10)  # you can pass other arguments to the log function to change the base
?log
exp(1)

# scientific notation works!
5000000000 * 1000
5e9 * 1e3
```

Think of a mathematical operation you need - can you guess how to do it in R?

# Assigning values to R objects

A key action in R is to store values in the form of R objects, and to examine the value of R objects.

```{r assigning_values}
val <- 3
val
print(val)

Val <- 7 # case-sensitive!
print(c(val, Val))

val <- 9 # can be overwritten
print(c(val, Val))

val = 6 # equal sign *can* be used for assignment but is not conventional
val

```
You can see these values change in the "Environment" tab in Rstudio.

We can work with (and store) sequences and repetitions
```{r sequences}
mySeq <- 1:6
mySeq

years <- seq(1952, 2007, by = 5)
years
length(years)

country <- rep("Afghanistan", 12)
country 

## This is a comment: here is an example of non-numeric data
countries <- c("Afghanistan", "India", "Pakistan")
countries
```

If we don't assign the output of a command to an object, we haven't saved it for later use.

R gives us a lot of flexibility (within certain rules) for assigning to (parts of) objects from (parts of) other objects.

# Be Lazy

If you're starting to type something you've typed before, or the long name of an R object or function, STOP!  You likely don't need to type all of that.

- Tab completion
- Command history 
    * up/down arrows
    * Ctrl-{up arrow} or Command-{up arrow}
- RStudio: select a line or block for execution
- Write re-usable code! Put your code in a file and use `source()`. For example: `source('myRcodeFile.R')`. Or, create functions you can use within an analysis.


# Vectors in R

The most basic form of an R object is a vector. In fact, individual (scalar) values are vectors of length one. The variables we created earlier (val, Val, mySeq, years) are all vectors!

We can concatenate values into a vector with `c()`.

```{r types_of_vectors}
## 1) numeric vector (also known as "double")
nums <- c(1.1, 3, -5.7)
myvals <- c(val, Val)
devs <- rnorm(5)
devs

## 2) integer vector
ints <- c(1L, 5L, -3L) # force storage as integer not decimal number
## 'L' is for 'long integer' (historical)
notints <- c(1, 5, -3)
identical(ints, notints)


## 3) character vector
chars <- c("pine", "apple")
paste(chars, collapse = '')
paste("character", "vector", sep = " ")

chars <- c('hi', 'hallo', "mother's", 'father\'s', 
   "She said, 'hi'", "He said, \"hi\"" )  
# the backslash is an "escape character"
# make sure opening and closing characters match!
chars



## 4) logical vector (or boolean vector)
bools <- c(TRUE, FALSE, TRUE)
bools

## combination
combi <- c("pizza", 9, TRUE)
combi

# How do I know what datatype something is?
typeof(bools)
```

# Working with indices and subsets

```{r indexing}
vals <- seq(2, 12, by = 2)
vals
vals[3]
```

Can you guess what these will do?
```{r indexing2}
vals[3:5]
vals[c(1, 3)]
vals[-c(1, 3)]
vals[c(rep(TRUE, 3), rep(FALSE, 2), TRUE)]
```

# Comparing Values
```{r comparisons}

vec1 <- c(1, 2, 3, 4, 5)
vec2 <- c(7, -4, 3, 4, 5.1)
vec1 < vec2
vec1 <= vec2

vec1 == vec2
vec1 != vec2
## careful: 
vec1 = vec2
identical(vec1, vec2)

## using 'or'
vec1 >= 3 | vec1 == exp(0)
## using 'and'
vec1 >= 3 & vec1 %% 2 == 0 
```

# Reading Data From A File

```{r load_data}

#######################################################################
## IMPORTANT: read in the Gapminder dataset from disk;
## first make sure your working directory is the 'modules' directory
getwd()
## if the result is not the 'modules' subdirectory of the bootcamp
## directory, set the working directly along the lines of this:
##
## setwd('/Users/jasminehughes/Documents/Intro_to_R/R_for_pharmacometricians/Modules')
##
## replace '/Users/jasminehughes/Documents/Intro_to_R/' with whatever directory you put the bootcamp
## materials in; e.g. on Windows it might be something like
## 'C:\\Users\\sarah\\R_for_pharmacometricians\\modules'
##
## If you've done that correctly, then the next command reads
## in the dataset from the 'data' directory. In the next
## command R finds that directory relative to the current
## working directory.
gap <- read.csv("../data/gapminder-FiveYearData.csv",
          stringsAsFactors = FALSE)
#######################################################################
```
Our data was imported as a dataframe. Dataframes are tabular: defined by rows and columns. When doing statistics & data analysis in R, you'll mostly be working with dataframes.

```{r looking_at_your_data}
## Look at part of the data
head(gap) # first few lines)
tail(gap) # last few lines
View(gap) # the whole shebang (avoid with very large datasets)
dim(gap)

## create a simple vector from the Gapminder dataset
gdp <- gap$gdpPercap
gdp[1:10]

## there's multiple ways to refer to a column
gap$gdpPercap[1:10]       #this method is probably the easiest, most legible
gap[[6]][1:10] 
gap[["gdpPercap"]][1:10] 

## you can index along the rows and columns
gap[5,1]

```
We can substitute values into vectors.
```{r subsituting_values}
gdp[4] <- 822
gdp[1:10]

vals <- rnorm(100)
vals[1:2] <- 0

## How does R process these next subset operations?
vals[vals < 0] <- 0
vals[1:8]

wealthy <- gdp[gdp > 100000]
gdp[1:10]
wealthy

# hint:
gdp_compared <- gdp > 100000
gdp_compared[1:10]
```

# Vectorized calculations

At the core of R is the idea of doing calculations on entire vectors.

```{r vectorized_calculations}
gdpTotal <- gap$gdpPercap * gap$pop

tmp <- gdpTotal[gap$year == "2007"]
gdpSubset <- tmp[1:20]
gdpSubset >= 1e6  

```


# R is a functional language

* Operations are carried out with functions. Functions take objects as inputs and return objects as outputs. 
* An analysis can be considered a pipeline of function calls, with output from a function used later in a subsequent operation as input to another function.
* Functions themselves are objects: 
```{r functions1}
median
class(median)
median(gap$lifeExp)
?median
```

Functions generally take arguments, some of which are often optional:
```{r functions2}
maxYear <- max(gap$year)
lifeExp <- gap$lifeExp[gap$year == maxYear]
median(lifeExp)
```

* We can embed function calls: 
```{r, hist, fig.cap = ""}
hist(rnorm(1000))
```

# Getting help about a function

To get information about a function you know exists, use `help` or `?`, e.g., `?lm`. For information on a general topic, use `apropos`.

```{r eval = FALSE}
help(lm)
?lm

?log

apropos("mean")
```

# Basic kinds of R objects

Vectors are not the only kinds of R objects.

### Vectors 

Vectors of various types (numeric (i.e., decimal/floating point/double), integer, boolean, character), all items must be of the same type

### Data frames

Collections of columns of potentially different types

```{r things_to_do_with_dfs}
head(gap)
dim(gap)
nrow(gap)
names(gap)
class(gap)
is.matrix(gap)
class(gap$year)
class(gap$lifeExp)
class(gap$country)
```

### Lists

Collections of disparate or complicated objects

```{r lists_get_complicated}
myList <- list(stuff = 3, mat = matrix(1:4, nrow = 2), 
   moreStuff = c("china", "japan"), list(5, "bear"))
myList
myList[[1]] # result is not (usually) a list (unless you have nested lists)
identical(myList[[1]], myList$stuff)
myList$moreStuff[2]
myList[[4]][[2]]
myList[1:3] # subset of a list is a list
myList$newOne <- 'more weird stuff'
names(myList)
```

Lists can be used as vectors of complicated objects. E.g., suppose you have a linear regression for each value of a stratifying variable. You could have a list of regression fits. Each regression fit will itself be a list, so you'll have a list of lists.


# Converting between different types of objects

You can use the `as()` family of functions.

```{r converting_between_types}
ints <- 1:10
as.character(ints)
as.numeric(c('3.7', '4.8'))
```

Be careful: R tries to be helpful and convert between types/classes when it thinks it's a good idea. Sometimes it is overly optimistic. 

```{r}
indices <- c(1.7, 2.3)
ints[indices]
ints[0.999999999]
```


# A (tiny) bit on plotting

R has several different plotting systems:

- *base* graphics
- *lattice* graphics
- *ggplot2* (an add-on package)

We'll see a little bit of *base* graphics here and then *lattice* and *ggplot2* tomorrow in Module 8.

```{r basic_plots, fig.cap =""}
hist(gap$lifeExp)
plot(gap$lifeExp ~ gap$gdpPercap)
boxplot(gap$lifeExp ~ gap$year)
```

# Graphics options

Check out `help(par)` for various [graphics settings](http://xkcd.com/833/); these are set via `par()` or within the specific graphics command (some can be set in either place), e.g.,
```{r, parstuff, fig.cap = ""}
par(pch = 16)
plot(gap$lifeExp ~ gap$gdpPercap, xlab = 'GDP per capita (dollars)',
   ylab = 'life expectancy (years)', log = 'x')
```

# Breakout

In general, your answers to any questions should involve writing code to manipulate objects. For example, if I ask you to find the maximum flight delay, do not scan through all the values and find it by eye. Use R to do the calculations and print results.

You can write your answers right below the questions by inserting a new chunk!

### Basics

1) Create a variable called 'x' that contains the mean life expectancy.

2) Use functions in R to round 'x' to two decimal places and to two significant digits.

3) Create a vector of GDP per capita in units of Euros rather than dollars.

4) Create a boolean (TRUE/FALSE) vector indicating whether total country GDP is greater than 1 trillion dollars. When entering 1 trillion, use R's scientific notation.

### Using the ideas

5) Use the boolean vector from problem 4 to produce a new dataset containing the data only from the biggest economies.
