---
title: 'Module 8: Pharmacokinetic R Packages'
author: "Jasmine Hughes"
date: "9/18/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# PKPDsim

>PKPDsim is an R package for numerical integration of ODE systems, in particular pharmacokinetic-pharmacodynamic (PK-PD) mixed-effects models.

>In pharmacometrics, models are often defined as systems of ordinary differential equations (ODEs). Although solving ODEs numerically in R is relatively straightforward using e.g. the deSolve library, the implementation of e.g. infusions and complex dosing regimens as well as the incorporation of random effects is cumbersome. Outside of R, a tool like Berkeley Madonna provides excellent interactivity features and is fast, but is much inferior to R regarding plotting functionality, cumbersome regarding implementation of dose regimens and multi-level variability, and also not open source.

>For fast numerical integration of the ODEs, the module uses the Boost C++ library under the hood.

```{r PKPDsim}
#devtools::install_github('InsightRX/PKPDsim')
library(PKPDsim)

# Get a pre-specified generic model
pk1 <- new_ode_model(model = "pk_2cmt_iv_auc")
pk1
new_ode_model() # To see available models

# Or specify your own system of ODEs
pk2 <- new_ode_model(
  code = "
  CLi = WT/70
  KEL = CLi/V
  dAdt[1] = -KA * A[1]
  dAdt[2] = +KA * A[1] -(CL*(WT/70)/V) * A[2]
  ", 
  declare_variables = c("KEL", "CLi"), 
  covariates = c("WT"))


```

To simulate time courses, you need to create a few more specialized objects to fully define the system.

```{r PKPDsim2}

# Define the regimen
reg1 <- new_regimen(
  amt = 1000,
  interval = 12,
  n = 6
)

# Define patient characteristics
pars1 <- list(
  "KA" = 1,
  "CL" = 5,
  "V" = 50,
  "Q" = 2.28,
  "V2" = 60
)

# Simulate the PK curve
pk1_res1 <- sim_ode(
  ode = pk1,              
  parameters = pars1,  
  regimen = reg1         
)

pk1_res1 %>%
  mutate(comp = case_when(
    comp == 1 ~ 'First Compartment',
    comp == 2 ~ 'Second Compartment',
    comp == 3 ~ 'AUC',
    TRUE      ~ 'Observation Compartment (TDMs)'
  )) %>%
  ggplot() +
  aes(x = t, y = y) +
  geom_line() +
  facet_wrap(~comp, scales = 'free')

```

Allows for exploration of different regimen on PK time curves, given a set of PK parameters, or vice versa.

```{r comparing_regimen}
reg2 <- new_regimen(
  amt = 750,
  interval = 8,
  n = 9
)

pk1_res2 <- sim_ode(
  ode = pk1,              
  parameters = pars1,  
  regimen = reg2         
)

bind_rows(mutate(pk1_res1, regi = 1),
          mutate(pk1_res2, regi = 2)) %>%
  mutate(comp = case_when(
    comp == 1 ~ 'First Compartment',
    comp == 2 ~ 'Second Compartment',
    comp == 3 ~ 'AUC',
    TRUE      ~ 'Observation Compartment (TDMs)'
  )) %>%
  ggplot() +
  aes(x = t, y = y, color = factor(regi)) +
  geom_line() +
  facet_wrap(~comp, scales = 'free')

```
```{r comparing_parameters}

pars2 <- list(
  "KA" = 1,
  "CL" = 8,      #60% higher clearance than pars1
  "V" = 50,
  "Q" = 2.28,
  "V2" = 60
)

pk1_res3 <- sim_ode(
  ode = pk1,              
  parameters = pars2,  
  regimen = reg1         
)

bind_rows(mutate(pk1_res1, par = 1),
          mutate(pk1_res3, par = 2)) %>%
  mutate(comp = case_when(
    comp == 1 ~ 'First Compartment',
    comp == 2 ~ 'Second Compartment',
    comp == 3 ~ 'AUC',
    TRUE      ~ 'Observation Compartment (TDMs)'
  )) %>%
  ggplot() +
  aes(x = t, y = y, color = factor(par)) +
  geom_line() +
  facet_wrap(~comp, scales = 'free')


```

A key part of NLMEM is interindividual/residual variability. These parameters can be specified through the `omega` and `ruv` arguments.

```{r variability}
pk1_omega <- sim_ode(
  ode = pk1,              
  parameters = pars2,  
  regimen = reg1,
  omega = c(0.2, 
            0.05, 0.1), #default is to assume exponential distribution
  n = 50 # do 50 simulations, sampling randomly
)

pk1_omega %>% 
  filter(comp == 'obs') %>%
  ggplot() +
  aes(x = t, y = y, color = as.factor(id)) +
  geom_line(alpha = 0.5) +
  theme(legend.position = 'none') 


```

```{r variability2}

pk1_ruv <- sim_ode(
  ode = pk1,              
  parameters = pars2,  
  regimen = reg1,
  omega = c(0.2, 
            0.05, 0.1), 
  n = 50,
  res_var = list(prop = 0.05, add = 0.1)  #5% proportional, 0.1 additive
)

pk1_ruv %>% 
  filter(comp == 'obs') %>%
  ggplot() +
  aes(x = t, y = y, color = as.factor(id)) +
  geom_line(alpha = 0.5) +
  theme(legend.position = 'none') 

```
More information available in function documentation, and at `http://pkpdsim.ronkeizer.com/`.

#PKNCA
> PKNCA provides functions to complete noncompartmental analysis (NCA) for pharmacokinetic (PK) data. Its intent is to provide a complete R-based solution enabling data provenance for NCA including tracking data cleaning, enabling calculations, exporting results, and reporting. The library is designed to give a reasonable answer without user intervention (load, calculate, and summarize), but it allows the user to override the automatic selections at any point.

PKNCA imports dplyr & tidyr, and provides some functionality for data cleaning.

```{r PKNCA_setup}
#install.packages("PKNCA")
library(PKNCA)

```

Similar steps to above, except that no model is necessary to specify.

- create regimen object

- create patient object

- create concentration object

- estimate AUC

```{r PKNCA_data}

conc_data <- data.frame(Subject = rep(1, 11),
                        Wt = rep(79.6, 11),
                        Dose = rep(4.02, 11),
                        Time = c(0, 0.25, 0.57, 1.12, 2.02, 3.82, 5.1, 7.03, 9.05, 12.12, 24.37),
                        conc = c(0.74, 2.84, 6.57, 10.5, 9.66, 8.58, 8.36, 7.47, 6.89, 5.94, 3.28))
# Let's take a peak at what this looks like first
conc_data %>%
  ggplot() +
  aes(x = Time, y = conc) +
  geom_line() +
  geom_point()
```


```{r PKNCA_object_creation}

dose_data <- filter(conc_data, Time == 0)


## Create a concentration object specifying the concentration, time, and
## subject columns.  
conc_obj <- PKNCAconc(conc_data,
                      conc ~ Time | Subject)
conc_obj

## Create a dosing object specifying the dose, time, and subject
## columns.  (Column names and data grouping must be the same between
## the two objects.)
dose_obj <- PKNCAdose(dose_data,
                      Dose ~ Time | Subject)
dose_obj

## Combine the concentration and dosing information
data_obj <- PKNCAdata(conc_obj, dose_obj)

## Calculate the NCA parameters
results_obj <- pk.nca(data_obj)

## Summarize the results
summary(results_obj)
```

PKNCA is doing a lot under the hood, and has a philosophy of making it easy to get started (like above) but also very flexible. Let's see what some of its default assumptions are:

```{r PKNCA_under_the_hood}
# (might be easier to read console output rather than RMarkdown output)
PKNCA.options()

# Can also use this function to change calculation specification
PKNCA.options(allow.tmax.in.half.life = TRUE)

```

See more options here: 
https://cran.r-project.org/web/packages/PKNCA/vignettes/Introduction-and-Usage.html

# Take-home?

The ability to create, share/collaborate and iterate on code/packages is one of the pro's of using R. See more PK packages here: 

- https://cran.r-project.org/web/views/Pharmacokinetics.html
- See VPC, which plots NONMEM outputs using ggplot: https://cran.r-project.org/web/packages/vpc/index.html 

Packages can be thought of a related set of functions that work together to accomplish a particular type of task.

Don't reinvent the wheel: if a package to do 90% of what you need to do exists, use it! You can write the remaining pieces yourself (and maybe contribute it to improve the original package!).

Don't Repeat Yourself (DRY): if you find yourself copy/pasting code a few times, write a function to handle it. (The opposite of DRY is WET (Write Everything Twice)) If you find yourself relying on a few related functions, turn it into a package. More on this in the next module....

# Breakout

### Basics

1. Use the data available in `datasets::Theoph` to estimate AUC for all the patients in the study, using non-compartmental analysis. Graphically represent your results. 

2. Use PKPDsim to simulate PK curves for a patient with less consistent dosing, by providing `times = c(...)` and `amt = c(...)` as arguments to `new_regimen()`.
