---
title: 'Module 2: Working with Data Objects'
author: "Jasmine Hughes"
date: "9/17/2019"
output: 
  slidy_presentation: default
---

```{r chunksetup, include=FALSE, cache=TRUE} 

if(!('Modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('Modules')
gap <- read.csv(file.path('..', 'Data', 'gapminder-FiveYearData.csv'), stringsAsFactors = FALSE)
```

# Matrices and Arrays

If you need to do numeric calculations on an entire non-vector object (dimension > 1), you generally want to use matrices and arrays, not data frames.

- Matrices are two-dimensional collections of values of the same type
- We can have numeric, integer, character, or logical matrices, character matrices.
- You can't mix types within a matrix

```{r matrices}
mat <- matrix(rnorm(12), nrow = 3, ncol = 4)
mat

# vectorized calcs work with matrices too
mat*4
mat <- cbind(mat, 1:3)
mat

# Let's convert the gapminder dataframe to a matrix:
gm_mat <- as.matrix(gap[ , c('lifeExp', 'gdpPercap')])
head(gm_mat)

# What do you think will happen?
gm_mat <- as.matrix(gap)
head(gm_mat)
```

Arrays are like matrices but can have more or fewer than two dimensions.
```{r}
arr <- array(rnorm(12), c(2, 3, 4))
arr
```

# Attributes

Objects have *attributes*.

```{r attr}
attributes(mat)

# Let's add an attribute
rownames(mat) <- c('first', 'middle', 'last')
mat
attributes(mat)

colnames(mat) <- c('A', 'B', 'C', 'D', 'E')
mat
attributes(mat)

names(attributes(gap))
attributes(gap)$names
attributes(gap)$row.names[1:10]
```

Now let's do a bit of manipulation and see if you can infer how R represents matrices internally.

```{r what_is_a_matrix}
mat[4]
attributes(mat) <- NULL
mat
is.matrix(mat)
```

**Question**: What can you infer about what a matrix is in R?

**Question**: What kind of object are the attributes themselves? How do I check?


# Missing values and other special values

Since it was designed by statisticians, R handles missing values very well relative to other languages.

* `NA` is a missing value
```{r}
vec <- rnorm(12)
vec[c(3, 5)] <- NA
vec
length(vec)
sum(vec)
sum(vec, na.rm = TRUE)
hist(vec)
is.na(vec)
```
Be careful because many R functions won't warn you that they are ignoring the missing values.

* To infinity and beyond
```{r}
big <- 1e500 
big
big + 7
```

* `NaN` stands for Not a Number
```{r}
sqrt(-5)
big - big
big
1/0
```

* `NULL`
```{r nulldata}
vec
vec <- c(vec, NULL) 
vec
length(vec)
a <- NULL
a + 7
a[3, 4]
is.null(a)
myList <- list(a = 7, b = 5)
myList$a <- NULL  # works for data frames too
myList
```

`NA` can hold a place but `NULL` cannot.
`NULL` is useful for having a function argument default to 'nothing'.


# Logical vectors
```{r}
gap2007 <- gap[gap$year == 2007, ]

wealthy <- gap2007$gdpPercap > 35000
healthy <- gap2007$lifeExp > 75

head(wealthy)
table(wealthy)

# note the vectorized boolean arithmetic
gap2007[wealthy & healthy, ]
gap2007[wealthy | healthy, ]
gap2007[wealthy & !healthy, ]

# what am I doing here?
sum(healthy)
mean(healthy)
```

**Question**: What do you think R is doing to do arithmetic on logical vectors?

# Data frames

A review from Module 1...

- Data frames are combinations of vectors of the same length, but can be of different types
- Data frames are what is used for standard rectangular (record by field) datasets, similar to a spreadsheet
- Data frames are a functionality that both sets R aside from some languages (e.g., Matlab) and provides functionality similar to some statistical packages (e.g., Stata, SAS)

```{r}
class(gap)
head(gap)
nrow(gap)
str(gap) # structure is abbreviated as 'str'
```

# Data frames are (special) lists!

```{r dataframes_and_lists}
is.list(gap)
length(gap)
gap[[3]][1:5]
lapply(gap, class) 
```

`lapply()` is a function used on lists; it works here to apply the `class()` function to each element of the list, which in this case is each field/column.

# But lists are also vectors!

```{r vectors_and_lists}
length(gap)
someFields <- gap[c(3,5)]
head(someFields)
identical(gap[c(3,5)], gap[ , c(3,5)])
```

In general the placement of commas in R is crucial, but here, two different operations give the same result because of the underlying structure of data frames.

# Factors
- A factor is a special data type in R used for categorical data. In some cases it works like magic and in others it is incredibly frustrating. 

```{r fac}
## let's read the Gapminder data in using the defaults for `read.csv`
gap <- read.csv(file.path('..', 'data', 'gapminder-FiveYearData.csv'))
class(gap$continent)
head(gap$continent) # What order are the factors in?
levels(gap$continent) 
summary(gap$continent)
```

- What if we don't like the order these are in? Factor order is important for all kinds of things like plotting, analysis of variance, regression output, and more

```{r pre_vs_post_factor}
# Does this factor order reflect what you'd want it to be?
factor(c("Pre-Intervention", "Post-Intervention"))
```

# Ordering the Factor
- Ordered factors simply have an additional attribute explaining the order of the levels of a factor
- This is a useful shortcut when we want to preserve some of the meaning provided by the order
- Think ordinal data

```{r orderedfac, fig.cap = ""}
pt_data <- data.frame(MRN = rep(c(97, 107, 402),2),
                      measurement_time = c(rep("Pre-Intervention", 3), 
                                           rep("Post-Intervention", 3)),
                      measurement_value = c(50, 65, 47, 104, 82, 91))
pt_data

boxplot(measurement_value ~ measurement_time, data = pt_data)

pt_data$measurement_time <- factor(pt_data$measurement_time, 
                                   levels = c("Pre-Intervention", "Post-Intervention"))
boxplot(measurement_value ~ measurement_time, data = pt_data)

```

# Reclassifying Factors
- Turning factors into other data types can be tricky. All factors have an underlying numeric structure.

```{r fac2}

levels(pt_data$measurement_time)
unclass(pt_data$measurement_time)
as.numeric(pt_data$measurement_time)
```

- Hmmm, what happened?
- Be careful! The best way to convert a factor is to convert it to a character first.
- This is a common source of bugs!

```{r}
students <- factor(c('basic','proficient','advanced','basic', 
      'advanced', 'minimal'))
score <- c(minimal = 3, basic = 1, advanced = 13, proficient = 7) # a named vector
score["advanced"]  # look up by name
students[3]
score[students[3]]
score[as.character(students[3])]
```

What went wrong and how did we fix it?  Notice how easily this could be a big bug in your code.

# Subsetting

There are many ways to select subsets in R. The syntax below is useful for vectors, matrices, data frames, arrays and lists.

```{r}
vec <- gap2007$lifeExp
mat <- matrix(1:20, 4, 5)
rownames(mat) <- letters[1:4]
mat
```
1) by direct indexing

```{r}
vec[c(3, 5, 12:14)]
vec[-c(3,5)]
gap[c(2,4), 5]
gap[c(2,4), 'lifeExp']

```


2) by a vector of logicals

```{r}
wealthy <- gap$gdpPercap > 50000
gap$gdpPercap[wealthy]
gap[wealthy, ]
```

What happened in the last subsetting operation?

3) by a vector of names
```{r}
mat[c('a', 'd', 'a'), ]
```

4) using *subset()*

```{r}
subset(gap, gdpPercap > 50000)
```

5) using *dplyr* tools such as *filter()* and *select()* -- more in an upcoming module!

# Assignment into subsets

We can assign into subsets by using similar syntax, as we saw with vectors.


```{r}
vec <- rnorm(20)
vec[c(3, 5, 12:14)] <- 1:5
vec
mat[2, 3:5] <- rnorm(3)
mat[mat[,1] > 0, ] <- -Inf
mat
```

Create an example of code where the number of elements you are trying to replace does not match the length vector you are trying to assign. Does the error message seem clear? Can you fix your (intentionally wrong) code?

# Strings

R has lots of functionality for character strings. Usually these are stored as vectors of strings, each with arbitrary length.

```{r}
chars <- c('hi', 'hallo', "mother's", 'father\'s', "He said, \"hi\"" )
length(chars)
nchar(chars)
paste("bill", "clinton", sep = " ")  # paste together a set of strings
paste(chars, collapse = ' ')  # paste together things from a vector

strsplit("This is the R bootcamp", split = " ")

countries <- as.character(gap2007$country)
substring(countries, 1, 3)
tmp <- countries
substring(tmp, 5, 10) <- "______"
tmp[1:20]
```
We can search for patterns in character vectors and replace patterns (both vectorized!)
```{r}
indexes <- grep("Korea", countries)
indexes
countries[indexes]
countries2 <- gsub("Korea Dem. Rep.", "North Korea", countries)
countries2[indexes]
```

# Regular expressions (regex or regexp)

Some of you may be familiar with using *regular expressions*, which is functionality for doing sophisticated pattern matching and replacement with strings. *Python* and *Perl* are both used extensively for such text manipulation. 

R has a full set of regular expression capabilities available through the *grep()*, *gregexpr()*, and *gsub()* functions (among others - many R functions will work with regular expressions). A particularly nice way to make use of this functionality is to use the *stringr* package, which is more user-friendly than directly using the core R functions.

You can basically do any regular expression/string manipulations in R, though the syntax may be a bit clunky at times.

# It might seem like a lot to remember
But you now have most of the basic building blocks. 
This "cheat sheet" might be helpful to refer to when attempting the questions below:
https://www.rstudio.com/wp-content/uploads/2016/10/r-cheat-sheet-3.pdf

# Breakout

### Basics

1) Extract the 5th row from the gapminder dataset.

2) Extract the last row from the gapminder dataset.

3) Count the number of gdpPercap values greater than 50000 in the gapminder dataset.

4) Set all of the gdpPercap values greater than 50000 to NA. You should probably first copy the `gap` object and work on the copy so that the dataset is unchanged (or just read the data into R again afterwards to get a clean copy).

5) Consider the first row of the gapminder dataset, which has Afghanistan for 1952. How do I create a string "Afghanistan-1952" using `gap$country[1]` and `gap$year[1]`? 

### Using the ideas

6) Create a character string using `paste()` that tells the user how many rows there are in the data frame - do this programmatically such that it would work for any data frame regardless of how many rows it has. The result should look like this: "There are 1704 rows in the dataset"

7) If you didn't do it this way already, extract the last row from the gapminder dataset without typing the number '1704'.

8) Create a boolean vector indicating if the life expectancy is greater than 75 and the gdpPercap is less than 10000 and calculate the proportion of all the records these represent.

9) Use that vector to create a new data frame that is a subset of the original data frame.


### Advanced

10) Create row names for the data frame based on concatenating the Continent, Country, and Year fields.
